---
title: "discussion1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Rmarkdown file of our first discussion:


The Chapman&Hall textbook Time Series: A Data Analysis Approach Using R and the Springer textbook Time Series Analysis and Its Applications has a companion R package astsa which contains time series data illustrated. https://github.com/nickpoison/astsa

Install it and get the data.

```{r , echo=FALSE}
# In case you haven't install devtools




```
For example, eqexp is:

```{r example , }
library("aTSA")
?astsa::eqexp

head(astsa::eqexp)
```
eqexp is a dataframe from the atsa package:
```{r runit , }
str(astsa::eqexp)
```


# **Plotting Time Series**
```{r runit2}
p_index = 1:1024
time_series_name = c("EQ6", "EX4")
time_series_to_be_plotted = astsa::eqexp[p_index, time_series_name] #row vector,column vector
plot.ts(time_series_to_be_plotted,
        plot.type = "multiple",
        nc = 1,
        main = "P waves") #explicit plot.ts call


```

now we are going to plot the data over a different time period
```{r runit3}
s_index = p_index+1024
colors_name = c("red", "blue")#red is p waves, blue is s waves (I believe)
time_series_to_be_plotted = astsa::eqexp[s_index, time_series_name]#time period/index,columns of TS data
plot.ts(time_series_to_be_plotted,
        plot.type="single",
        main="S waves",
        col=colors_name, #associate red with column EQ6, blue with column EX4
        ylab="somethings")
legend("bottomleft",
       time_series_name,
       col=colors_name,
       lty=1)
```


# **Simulating Time Series**
![](https://render.githubusercontent.com/render/math?math=s_t%20%3D%20%5Cbegin%7Bcases%7D%0A%20%20%20%20%20%200%2C%20%26%20t%3D1%2C%5Cdots%2C125%20%5C%5C%0A%20%20%20%20%20%2010%5C%3Be%5E%7B-%5Cfrac%7Bt-125%7D%7B25%7D%7D%5Ccos(2%5Cpi%20t%2F4)%20%2C%26%20t%3D126%2C%5Cdots%2C250%0A%20%20%20%20%20%20%5Cend%7Bcases%7D)

Now, a Signal with White Noise example:
```{r runit4}
s_fun = function(t) ifelse(t<=125, 0, 10*exp(-(t-125)/25)*cos(2*pi*t/4))
omega_fun = function(t) rnorm(length(t))

t_index = 1:250
set.seed(19460614)
x = ts(s_fun(t_index) + omega_fun(t_index))

str(x) #str function tells us the object type of the argument (x)
```
```{r runit5}
plot(x)
```


```{r runit6}
acf(x) #displays the autocorrelation function of x
```


there is clearly statistically significant autocorrelation here....
Ideally, the acf quickly decreases towards low autocorrelation and stays that way -> means stationarity and weak dependence assumptions are valid.

In this instance, the weak dependence assumption is not met
