---
title: "project1-temperature"
author: "Collin"
date: "10/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



### **Project 1**

File "temperature.csv" contains average monthly temperature, say xt, of a region with humid continental climate from January 1964 until December 1975. In this problem, we are going to fit a special type of structural model,

**a)** *(a) plot the data and elaborate on the general pattern of the data.*
```{r}
temperature <- read.csv("~/Google Drive/UC Davis/UC Davis/fall_quarter_2020/STA137-Time_Series/temperature.csv", sep="")
month = factor(cycle(temperature$temperature),12) #create the seasonal component


myts <- ts(temperature, start=c(1964, 1), end=c(1975, 12), frequency=12) 


#a
plot(myts)

```
The main thing worth mentioning is that there is clearly a seasonal component to this 
temperature time series!

**b** *Fit the regression model:*
```{r}
fit = lm(temperature ~ 0 + factor(cycle(myts)),data = myts,na.action = NULL) #no intercept by including 0
summary(fit)


#i
janToFeb = 20.650-16.608
febToMar = 32.475-20.650
marToApr = 46.525-32.475
aprToMay = 58.092 - 46.525
mayToJun = 67.500 - 58.092
juntoJuly = 71.717-67.500
julyToAug = 69.333 - 71.717
augToSep = 61.025-69.333
sepToOct = 50.975 - 61.025
octToNov = 36.650 -50.975
novToDec = 23.642-36.650

vecOfTempChanges = c(janToFeb,febToMar,marToApr,aprToMay,mayToJun,juntoJuly,julyToAug,augToSep,sepToOct,octToNov,
  novToDec)


tempChange = matrix(vecOfTempChanges,nrow = 1,ncol=11,byrow = TRUE)
colnames(tempChange)= c('janToFeb','febToMar','marToApr','aprToMay','mayToJun','juntoJuly','julyToAug','augToSep','sepToOct','octToNov',
                        'novToDec')
tempChange

#ii
fit = lm(temperature ~ 0 + factor(cycle(myts,12)),data = myts,na.action = NULL)
fittedTS = ts(fit$fitted.values,start=c(1964, 1), end=c(1975, 12), frequency=12) 

plot.ts(myts, xlab = "Time",main = "Observed (black) vs. Fitted Temp:1964-1975")
par(new=TRUE)
lines(fittedTS,col=3)

```
The fitted values appear to fit the time series exceptionally well.There is
significant overlap between the fitted and observed temperature values over the 
time series, and the residuals appear to be miniscule.



**c** *In this part we try to find dominant periodic components in the data. Fit the regression model*
```{r}
#fit the model
tIndex = 1:144
someFunction = function(t){
  cos((2*pi*1*t/12)) + sin((2*pi*1*t/12)) + cos((2*pi*2*t/12)) + sin((2*pi*2*t/12)) +
    cos((2*pi*3*t/12)) + sin((2*pi*3*t/12)) + cos((2*pi*4*t/12)) + sin((2*pi*4*t/12)) +
    cos((2*pi*5*t/12)) + sin((2*pi*5*t/12)) + cos((2*pi*6*t/12)) + sin((2*pi*6*t/12))
}

reg = lm(temperature$temperature ~ cos((2*pi*1*tIndex/12)) + sin((2*pi*1*tIndex/12)) + cos((2*pi*2*tIndex/12)) + sin((2*pi*2*tIndex/12)) +
           cos((2*pi*3*tIndex/12)) + sin((2*pi*3*tIndex/12)) + cos((2*pi*4*tIndex/12)) + sin((2*pi*4*tIndex/12)) +
           cos((2*pi*5*tIndex/12)) + sin((2*pi*5*tIndex/12)) + cos((2*pi*6*tIndex/12)) + sin((2*pi*6*tIndex/12)))


#plotting this function over 1-144 looks like this
Xt = ts(someFunction(tIndex),start=c(1964, 1), end=c(1975, 12), frequency=12) 
plot(Xt)


#(i) If the model is correct, determine the significant frequencies in the time series. Based on your analysis, how many months it takes for the time series to repeat itself.

#regressing temperature on this function yields:
summary(reg)
```

if this model is correct, there appear to be 5 statistically significant terms at the
5% significance level: the intercept, and the 1st, 2nd, 4th, and 6th periodic components.
It also appears to take about 12 months for the series to repeat itself. (This is
consistent with the earlier model we used.)



```{r}
#(ii)Graph the data, xt, and superimpose the fitted values, say x^t, on the graph.

fittedTS2 = ts(reg$fitted.values,start=c(1964, 1), end=c(1975, 12), frequency=12)
plot.ts(myts, xlab = "Time",main = "Observed (black) vs. Fitted:1964-1975")
par(new=TRUE)
lines(fittedTS2,col=2)

```


```{r}
#(iii) Examine the residuals, ð‘¥ð‘¡âˆ’ð‘¥Ì‚ð‘¡, and state your conclusions, Does it appear that the model fits the data well (do the residuals look white)? Does it appear that the residuals are a sample from a Normal distribution? Justify your answer.


hist(reg$residuals)

qqnorm(reg$residuals)
qqline(reg$residuals)



```

based on the overlaying of the fitted values onto the actual time series, the residuals
appear small, signifying that these periodic components explain much of the variation
in the model of temperature.

the residuals do in fact appear to be (approximately) normally distributed. While
not perfectly normal, a large proportion of the errors are centered around 0, and they
disperse to the left and right in a bell shape.
